tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-08 paper.tex
Updates

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of samples now required.
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

@@

\section{Phenotypic Model}

Complex traits are traits that are influenced by a mixture of genetic and
environmental factors. The additive model is a common assumption used in GWAS
\citep{uffelmann2021}, and \texttt{tstrait} assumes that the traits are
obtained from the following phenotypic model,
\begin{align}\label{eq:additive-model}
    y=X\beta+\epsilon,
\end{align}
where $\beta$ is the effect size, $X$ is the matrix that describes the number
of causal mutations in the individual, and $\epsilon$ is the environmental
noise. The environmental noise is simulated from the following normal
distribution,
\begin{align}\label{eq:env}
    \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
\end{align}
where $V_G$ is the additive genetic variance from the simulated effect sizes
and $h^2$ is the narrow-sense heritability that is defined by the user.

\section{Package Overview}

\subsection{Effect Size}

The first step of quantitative trait simulation in \texttt{tstrait} is to set
the trait model distribution by using \emph{trait\_model} function, as in the
following example:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
  distribution='normal', mean=0, var=1
)
\end{lstlisting}
The \texttt{tstrait} package supports effect size simulation from five
univariate distributions and one multivariate distribution, which are taken
from commonly used statistical distributions to simulate effect sizes
\citep{gaynor2021,haller2023}. The multivariate normal distribution trait model
is used to simulate multiple traits, assuming that they are pleiotropic. The
trait model is used as an input of \emph{sim\_trait} function to simulate
effect sizes:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
   ts, num_causal=5, model=model
)
\end{lstlisting}
\emph{ts} is the tree sequence input, and the user can set the number of causal
sites that is affecting the trait. The output will be a \texttt{pandas}
dataframe \citep{pandas} that describes the simulated effect sizes.

\subsection{Genetic Value}

Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
\begin{lstlisting}[language=Python]
genetic_result = tstrait.sim_genetic(
   ts, trait_df, alpha=0
)
\end{lstlisting}
\texttt{tstrait} offers flexibility in genetic value simulation, as users can
define their own effect size dataframe, in addition to using the simulated
output of \emph{sim\_trait} function. This allows users to use the output from
other simulators to simulate traits in \texttt{tstrait}.

This is the most computationally intensive part of the simulation algorithm, as
it would be necessary for the algorithm to count the number of causal mutations
in each individual inside the dataset. The \texttt{sim\_genetic} function uses
the tree-traversal algorithm to analyze the tree sequence data
\citep{ralph2020}, and uses the Python package \texttt{Numba} \citep{numba} to
speed up the code. This enables \texttt{tstrait} to be an efficient
quantitative trait simulator of a tree sequence data.

\texttt{tstrait} also supports effect size simulation from the frequency
dependent model, where the simulated effect sizes from the causal mutation
having allele frequency $p$ will be multiplied by $[2p(1-p)]^\alpha$. The
$\alpha$ parameter is a user-defined parameter, which is used to control the
degree of frequency dependence on simulated effect sizes. It has been suggested
that effect sizes from rarer variants can have increased magnitude from
negative selection, and the model above with a negative $\alpha$ value is used
to conduct realistic simulations \citep{speed2017}.

\subsection{Phenotype}

Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
\begin{lstlisting}[language=Python]
phenotype_df = tstrait.sim_env(
   genetic_result.genetic, h2=0.3, random_seed=1
)
\end{lstlisting}
The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).

After completing the simulation, \texttt{tstrait} returns two dataframes to
describe the simulated genetic effect sizes and phenotypes. The resulting
dataframes include information regarding the site IDs, individual IDs and
allele frequencies, so it would be possible to regenerate the simulated results
whenever required.

Alternatively, all the above steps of the simulation can be conducted by using
\emph{sim\_phenotype}:
\begin{lstlisting}[language=Python]
sim_result = tstrait.sim_phenotype(
  ts, num_causal=100, model=model, h2=0.3
)
\end{lstlisting}
The user only needs to input the number of causal sites, trait model and the
narrow-sense heritability to simulate quantitative traits. By using this
function, the user does not need to modify the details of the quantitative
trait simulation.

\section{Computational Time}

\texttt{tstrait} offers a very efficient way of simulating quantitative traits
based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales
with the number of individuals on simulated human genomes in \texttt{stdpopsim}
\citep{adrion2020}. The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

>>>>

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of samples now required.
[TODO discuss bespoke stuff like Martin et al]
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

@@

\section{Results}
\subsection{Phenotypic Model}
The additive model is a common assumption in GWAS
\citep{uffelmann2021}, and we assume that a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal mutations in the individual, and $\epsilon$ is the environmental
noise.
[TODO more discussion here of what this means in an ARG. Be clear about
causal mutations]
% Environmental noise is normally distributed with mean zero
% and a variance of $V_G(1-h^2)/{h^2}$,
% % distribution,
% % \begin{align}\label{eq:env}
%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% % \end{align}
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Effect Sizes}
The first step of quantitative trait simulation in \texttt{tstrait} is to
define the trait model distribution, as follows:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
    distribution='normal', mean=0, var=1)
\end{lstlisting}
\texttt{Tstrait} currently supports effect size simulation from five
commonly used univariate distributions~\citep{gaynor2021,haller2023}
and one multivariate distribution.
The multivariate normal distribution is used to simulate multiple
pleiotropic traits.
The trait model is used as an input of \texttt{sim\_trait} function to simulate
effect sizes, along with the input ARG and number of causal sites:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
    arg, num_causal=5, model=model)
\end{lstlisting}
The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
describing the simulated effect sizes and corresponding
causal site IDs.

\subsection{Genetic Value}
Given a set of causal sites and effect sizes defined in a dataframe
and the input ARG, the \emph{genetic values} [TODO DEFINE ME]
are simulated as follows:
\begin{lstlisting}[language=Python]
genetic_result = tstrait.sim_genetic(
   arg, trait_df, alpha=0)
\end{lstlisting}
The $\alpha$ parameter here is used to
supports effect size simulation from a frequency
dependent model, where the simulated effect sizes from the causal mutation
having allele frequency $p$ are multiplied by $\left(2p(1-p)\right)^\alpha$.
An $\alpha$ value of zero as in this example means there is
no frequency dependence. A negative value of $\alpha$
can be used to model the increased effect sizes on rare variants
resulting from negative selection~\citep{speed2017}.
[TODO is this correct? What's the interpretation of positive alpha?]

A major benefit of this modular approach to simulating effect sizes
and genetic values separately is the flexibility this offers
users. Because the causal sites and effect sizes are specified in a
simple tabular format, users can easily develop their own approach
to simulating these values. Alternatively, other simulators
such as SLiM~\citep{haller2023} that generate effect sizes and
causal mutations during the progress of a forwards-time simulation
(to model the effect of selection, for example) could output these
values to a CSV or similar file. The modularity here and simple
input data format is specifically intended to facilitate such
interoperability between simulators.

Generating the genetic values for a trait is by far the
most computationally intensive part of the simulation,
as we are essentially counting the number
of causal mutations in each individual. This is done by a
straightforward tree traversal algorithm in \texttt{tstrait},
accelerated using the \texttt{numba} JIT compiler for Python.
As we see in later sections, this yields very good performance
at the largest scale, without requiring any additional code in
compiled languages such as C.

\subsection{Phenotype}
The final phenotype simulation is done in \texttt{tstrait}
by imposing some environmental noise.
We assume environmental noise is normally distributed with mean zero
and a variance of
\[
V_G(1-h^2)/{h^2}
\]
where $V_G$ is the additive genetic variance from the simulated effect sizes,
and $h^2$ is the narrow-sense heritability parameter provided as input.
Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
and narrow-sense heritability as parameters:
\begin{lstlisting}[language=Python]
phenotype_df = tstrait.sim_env(
    genetic_result.genetic, h2=0.3)
\end{lstlisting}
This simulation returns a dataframe
describing the simulated genetic value, environmental noise and
resulting phenotype for each individual at each causal site.

As discussed above, there are significant advantages to providing a modular
API. For convenience, we also provide a single
\texttt{sim\_phenotype} function that combines the process of simulating
effect sizes, generating genetic values and imposing environmental noise.

\subsection{Performance}
[START HERE]
\texttt{tstrait} offers a very efficient way of simulating quantitative traits
based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales
with the number of individuals on simulated human genomes in \texttt{stdpopsim}
\citep{adrion2020}. The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.
